---
title: "Colorado COVID-19 Likelihood of Releases"
author: "Albert Sun"
date: "9/30/2020"
output:
  pdf_document: default
  html_document: default
---

```{r message=FALSE}
library(tidyverse)
library(patchwork)
library(janitor)
library(broom)
library(knitr)
library(here)
library(rms)
library(skimr)
```

```{r global options, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

# Background

## Motivation

We wish to assess the quality of data collection and efficacy of using existing open-source jail databases to understand how jail populations respond to outside legislation and large trends, particularly disasters like COVID-19. The Colorado Jail Database is currently the most comprehensive, open source published, and catalogued state-wide jail database in the United States. Thus, we seek to conduct various statistical analyses on Colorado’s jail database to understand how similar data collections can motivate statistical conclusions in the future. If proven insightful, Colorado can be an effective model that other states can look towards to adopt a similar system of data collection.

Colorado has a uniform system for collecting jail data in the state, known as the HB19-1297 Jail Data Collection. This dashboard is run by the Colorado Department of Corrections within the Division of Criminal Justice. HB19-1297 was a bipartisan bill primarily sponsored by State Representatives Michael Weissman (D) and Hugh McKean (R). HB19-1297 was signed May of 2019. HB19-1297 expanded the information that jail keepers must submit to the Division of Criminal Justice, mandating that unless data is submitted quarterly, jail keepers could be subject to a $500 fine. 

Some stakeholders in Colorado, and across the country, have attempted to reduce jail populations at the beginning of the outbreak. Colorado Governor Jared Polis signed an executive order relaxing the standards for early release in March in fear of the effects of excessive overcrowding in jails. Jails who initially thought of expanding have halted their plans due to general decreases in jail populations. Overall, Colorado has had a reported net decrease in jail population during the COVID-19 criss, which can immediately be seen on the Colorado dashboard itself. 

## Research Question 

What demographic or jail information is correlated with a county jail decreasing its jail population at the outbreak of and during the COVID-19 pandemic? 

We seek to impute demographic variables from other data sets to further understand this question.

## Data Comments

The public and published HB 19-1297 dataset represents 3 collection periods in 2020, split up into yearly quarters. Q1 was Jan-Mar, Q2 was in Apr-Jun, Q3 was in Jul-Aug.

The 23 columns represent variables: the quarter, the year, county jail, jail management system, etc.

The 2280 rows generally reflect specific jail information per each quarter; however, the reason why there are 2280 rows instead of 152 rows (the number of jails times three quarters in Colorado) is because each jail has 15 rows separated into different areas of measurement, i.e. "Number of inmates", "Sentenced", etc. 

# Data Preparation

## Data Loading

```{r read colorado and pop}
#load colorado hb-19 1279
colorado <- read_csv(here("data", "HB19-1297Data.csv")) %>%
  clean_names()

#load colorado population data
pop <- read_csv(here("data", "colorado-population.csv")) %>%
  clean_names() %>%
  filter(county != "Total") %>%
  mutate(county = str_sub(county, 2, -18))

#load colorado population demographics data
demo <- read_csv(here("data", "mit-demographics.csv")) %>%
  filter(state == "Colorado") 

demo <- demo %>%
  mutate(liberal = factor(if_else(clinton16 - trump16 > 0, 1, 0))) %>%
  mutate(urbanicity = factor(ruralurban_cc)) %>%
  mutate(urbanicity = fct_collapse(urbanicity, 
                      metro = c("1", "2", "3"), 
                      urban = c("4", "5", "6", "7"), 
                      rural = c("8", "9"))) %>%
  mutate(urbanicity = fct_relevel(urbanicity,
                                     'rural',
                                     'urban')) %>%
  select(county, lesscollege_whites_pct, black_pct, rural_pct, urbanicity, liberal)

demo
```


## Check Proportion Missing

Instead of leaving blank values in missing columns, the Colorado HB 19-1297 jail dataset uses the `not_available` column to annotate and comment on missingness. They add `0` to a datapoint that is missing. Thus, because we cannot use conventional functions like is.na() to detect missingness, we will take a look at the jail observations that contain missing data. Here are the 10 most common NA messages.

```{r}
colorado %>% 
  count(not_available) %>%
  group_by(not_available) %>%
  arrange(-n) %>%
  ungroup %>%
  slice(1:10)
```

Most of the data that is missing is that a jail's JMS (Jail Management System) might not break down types of sentences by gender, race, or ethnicity. When conducting data analysis on race and gender for some particular measures, it will be a good idea to remove these rows, or at least account for them. 

Out of 2280 rows, there are `r 2280 - 1792` (2280-1792) rows with some sort of `not_available` message. 

This means that `r 448 / 2280 * 100`% of the data has some sort of `not_available` message to it, which is relatively low.

## Check Class

Check class of data:

```{r glimpse at the data}
glimpse(colorado)
```
Judging from the datatypes above, we will change Qtr to become a factor variable, because Qtr represents periodic stages of data collection, not a continuous value.

```{r}
colorado <- colorado %>%
  mutate(qtr = factor(qtr))
```

Overall, the other variables seem to have the correct data type.

## Investigate Missingness

```{r}
colorado <- colorado %>% 
  mutate(isNA = !is.na(not_available)) 

colorado %>%
  count(qtr, isNA) %>%
  filter(isNA == TRUE)

colorado
```

Missingness was generally reduced throughout the three quarters of jail data collection in 2020, possibly suggesting improvements in jail collection throughout this time period.

## Merging Datasets

Let's focus on only quarter 1 and 3 as before and after end points for COVID-19. This will allow us to adjudicate the full response to the ongoing COVID-19 virus.

```{r}
colorado_num_inmates <- colorado %>%
  filter(measure == "Number of inmates") %>%
  filter(qtr == 1 | qtr == 3)

colorado_num_inmates
```

```{r}
colorado_num_inmates %>%
  count(county) %>%
  filter(n == 1)
```

Remove 4 jails that don't have both first and third quarter: Grand, Huerfano, Las Animas, Saguache. Remove unneeded variables. Merge unknown and other race.

```{r}
colorado_num_inmates <- colorado_num_inmates %>%
  filter(county != "Grand" & 
           county != "Huerfano" &
           county != "Las Animas" &
           county != "Saguache") %>%
  select(-c(not_available, isNA, jms, qtr_year, measure, deaths, 
            other_gender, 
            bookings, releases)) %>%
  mutate(other_race = unknown_race + other_race) %>%
  select(-c(unknown_race)) %>%
  arrange(county)

colorado_num_inmates
```

Pivot the dataset wider to treat each jail as its own observation, with the ability to create new columns representing the changes in population counts.

```{r pivot wider}
colorado_num_inmates <- colorado_num_inmates %>% 
  pivot_wider(names_from = qtr, values_from = capacity:unknown_ethnicity) 
```

Join the Colorado HB 19-1297, population, and MIT Election Lab Aggregated Demographic datasets.

```{r join the datasets}
colorado_num_inmates <- colorado_num_inmates %>%
  inner_join(pop, by = "county") %>%
  inner_join(demo, by = "county")
```

Create new dataset called colorado_num_percent, which transforms all the variables into percent changes in order to standardize the numbers across different sizes of jails. 

```{r standardizing pcts}
colorado_num_percent <- colorado_num_inmates %>%
  mutate(difference = (total_3 - total_1)/total_1) %>%
  mutate(jail_male_pct = male_1 / total_1) %>%
  mutate(jail_black_pct = black_1 / total_1) %>%
  mutate(jail_hispanic_pct = hispanic_1 / total_1) %>%
  select(county, difference, lesscollege_whites_pct, jail_male_pct, jail_black_pct, jail_hispanic_pct,
         population, rural_pct, black_pct, liberal, urbanicity)

colorado_num_percent
```

## Exploratory Data Analysis

General Summary Statistics:

```{r distribution of percent removed per jail}
skim(colorado_num_percent)
```

Based on the response variable `difference`, there has been an average 0.164% decrease in population per jail. The distribution is right skewed.

```{r}
demo %>%
  ggplot(aes(urbanicity)) + 
  geom_bar() + 
  labs(title = "Distribution of counties per urbanicity")
```

```{r}
colorado %>%
  mutate(jms = str_replace(jms, "Eforce", "E-Force")) %>%
  mutate(jms = str_replace(jms, "eForce", "E-Force")) %>%
  count(jms) %>%
  arrange(-n) %>%
  mutate(n = n / 5) %>%
  slice(1:5) %>%
  ggplot(aes(x = reorder(jms, -n), y = n)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Jail Management System",
       y = "Count",
       title = "Top 5 utilized jail management systems in Colorado")
```

The top utilized jail management system in Colorado are E-Force, Spillman, New World, Crimestar, and Tiburon.

```{r}
colorado_population <- colorado %>%
  arrange(qtr) %>%
  group_by(qtr) %>%
  summarise(total = sum(total)) 

colorado_population
```

As you can see, Colorado jail population has significantly reduced during COVID-19. 

```{r Jail Population Distribution}
# facet version

colorado_dist <- colorado %>%
  group_by(county, qtr) %>%
  summarise(total = sum(total))

colorado_dist %>%
  ggplot(aes(x = total)) + 
  geom_histogram(bins = 50) + 
  facet_grid(qtr ~ .) + 
  labs(title = "Distribution of Colorado Jail Size (by Population) in different
Quarters",
       y = "Count",
       x = "Jail Population")
```

The population per jail throughout Colorado is Unimodal, right-skewed distribution with significant outliers on the right of the
graph.

# Model Building

## Variables:

`liberal` (Factor): 1 if county majority voted for Clinton in 2016, 0 if county majority voted for Trump

`lesscollege_whites_pct` (Double): white population with an education of less than a bachelor's degree as a percentage of total population

`population` (Double): Total population of a county

`jail_male_pct` (Double): Percent male of a jail

`jail_black_pct` (Double): Percent Black of a jail

`jail_hispanic_pct` (Double): Percent Hispanic of a jail

`urbanicity` (Factor): metro - Counties in Metropolitan Areas. urban - Counties in Urban Areas. rural - Counties in Rural Areas. Definitions of Metropolitan, Urban, and Rural are designated by the US Office of Management and Budget (OMB) delineation as of February 2013
`
`black_pct` (Double): Percentage of Black people in general county population

## Initial Model Fitting

First, put as many predictor variables as possible to create a full model:

```{r model fit}
full <- lm(difference ~ liberal + 
     lesscollege_whites_pct + 
     population +
     jail_male_pct + 
     jail_black_pct + 
     jail_hispanic_pct+ 
     urbanicity + 
     black_pct,
   data = colorado_num_percent)

full %>%tidy
```

Backwards selection with AIC as selection criterion to choose the model with the best AIC.

```{r backwards selection}
int_only_model <- lm(difference ~ 1, data = colorado_num_percent)

covid_model <- step(full, scope = formula(int_only_model), direction = "backward")
```

Based on backwards AIC selection, the two significant predictors for are the percent of males in a jail and whether a jail is in a rural, urban, or metropolitan area.

## Interaction Term

Now, let's use nested F-test to check if the addition of an interaction term between the two variables above is statistically significant:

```{r nested F-test}
reduced_model <- covid_model
full_model <- lm(difference ~ 
     jail_male_pct + 
     urbanicity + 
     jail_male_pct * urbanicity,
   data = colorado_num_percent)

anova(reduced_model, full_model) %>%
  tidy() %>%
  kable(digits = 3)
```

Since F-statistic is high and p-value is close to 0, the interaction effect between jail_male_pct * ruralurban_cc is statistically significant. 

# Model Interpretations:

```{r}
full_model %>%
  tidy(conf.int = TRUE) %>%
  kable(digits = 3)
```

## Urbanicity: 

Rural jails have failed to slow to decreasing their jail population during COVID-19 in comparison to urban/metropolitan areas:

- A jail in an urban area is expected to decrease its population 12 percent more than a jail in a rural area, on average.

- A jail in an metropolitan area is expected to decrease its population 13 percent more than a jail in a rural area, on average.

## Male Population Percentage and its Interaction with Urbanicity: 

Jails in rural and urban areas with a higher male population have a higher chance to decreasing their jail population. Jails with higher male populations in metropolitan areas have a higher chance of increasing their jail population. Specifically, 

- For rural jails, for every one percent increase in male inmates, there is expected to be a 14 percent decrease in jail population between Jan to Sept 2020, on average.

- For urban jails, for every one percent increase in male inmates, there is expected to be a 1 percent decrease in jail population between Jan to Sept 2020, on average.

- For metropolitan jails, for every one percent increase in male inmates, there is expected to be a 1 percent increase in jail population between Jan to Sept 2020, on average. 

# Model Conditions:

## Check Conditions

```{r conditions}
model_aug <- augment(full_model) %>%
   mutate(obs_num = row_number()) #add row number to help with graphing

resid_fitted <- ggplot(data = model_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted values",
    y = "Residual",
    title = "Residuals vs. Predicted") 

resid_hist <- ggplot(data = model_aug, aes(x = .resid)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Residuals", title = "Dist. of Residuals") 

resid_qq <- ggplot(data = model_aug, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal QQ-plot of residuals") 

conditions_plot <- resid_fitted / (resid_hist + resid_qq)

conditions_plot
```

4 Conditions for Linear Regression:

1. Linearity: Partially Satisfied - Although there is no distinguishable pattern in the residuals vs. predicted plot, the points are generally scattered. There seem to be some high leverage points that are worth investigating. However, lee-way will be given given the small number of observations. Overall, the residuals are randomly scattered.

2. Normality: Satisfied - the points genearlly fall along a straight diagonal line on the normal quantile plot. 

3. Constant Variance: Satisfied -  the vertical spread of the residuals remains relatively constant across the plot.

4. Independence: Satisfied - The error for one county does not tell us anything about the error for another county. We also put urbanicity of the counties as one of the variables, which avoids some of the problems of spatial collinarity/correlation. However, in order to improve this model, we can add a "region" variable to account for the locations of the counties in the state of Colorado. 

## Model Diagnostics

**Cook's distance**

Let's investigate those possible high-leverage points:

```{r cooks d}
#scatterplot of cook's d vs obs num
ggplot(data = model_aug, aes(x = obs_num, y = .cooksd)) + 
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept = 1, color = "red") +
  geom_hline(linetype = "dotted", yintercept = 0.5, color = "red") +
  labs(x = "Observation Number", y = "Cook's D") +
  geom_text(aes(label=ifelse(.cooksd > 1,
                             as.character(obs_num), "")), nudge_x = 4)
```

Jackson County (Observation 23), which has a super small county jail, is a high leverage county. This is because it increased from having 2 people to 17 people in its jail over COVID-19. It is an influential point, meaning that it has a large impact on the coefficients and standard errors used for inference. 

Because the goal of the model is explanation as opposed to prediction, it is worth keeping this point in the model.


