---
title: "Colorado COVID-19 Likelihood of Releases"
author: "Albert Sun"
date: "9/30/2020"
output:
  pdf_document: default
  html_document: default
---

```{r message=FALSE}
library(tidyverse)
library(patchwork)
library(janitor)
library(broom)
library(knitr)
library(here)
```

```{r global options, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

# Data Loader

```{r read colorado and pop}
colorado <- read_csv(here("data", "HB19-1297Data.csv")) %>%
  clean_names()

pop <- read_csv(here("data", "colorado-population.csv")) %>%
  clean_names() %>%
  filter(county != "Total") %>%
  mutate(county = str_sub(county, 2, -18))
```

```{r read demographics}
demo <- read_csv(here("data", "mit-demographics.csv")) %>%
  filter(state == "Colorado") 

demo <- demo %>%
  mutate(liberal = factor(if_else(clinton16 - trump16 > 0, 1, 0))) %>%
  mutate(ruralurban_cc = factor(ruralurban_cc)) %>%
  mutate(ruralurban_cc = fct_collapse(ruralurban_cc, 
                      metro = c("1", "2", "3"), 
                      urban = c("4", "5", "6", "7"), 
                      rural = c("8", "9"))) %>%
  mutate(ruralurban_cc = fct_relevel(ruralurban_cc,
                                     'rural',
                                     'urban')) %>%
  select(county, lesscollege_whites_pct, black_pct, rural_pct, ruralurban_cc, liberal)

demo
```


# Overall Comments

Overall, the HB 19-1297 data set is the only statewide jail database that offers an open source ".csv" file for the public to see. Even other states who do collect jail population data, like California and Texas, don't have their data published completely open source; instead, researchers are forced to scrape the data periodically from their website. As such, we find that the HB-1297 data set to be the most reproducible thus far.

The 23 columns represent variables: the quarter, the year, county jail, jail management system, etc.

The 2280 rows generally reflect specific jail information per each quarter; however, the reason why there are 2280 rows instead of 152 rows (the number of jails times three quarters in Colorado) is because each jail has 15 rows separated into different areas of measurement, i.e. "Number of inmates", "Sentenced", etc. In more-technical SQL terms, it seems like the measure column was cross joined with jail column. 

# 1. Check Proportion Missing

Instead of leaving blank values in missing columns, the Colorado HB 19-1297 jail dataset uses the `not_available` column to annotate and comment on missingness. They add `0` to a datapoint that is missing. Thus, because we cannot use conventional functions like is.na() to detect missingness, we will take a look at the jail observations that contain missing data. 

Here are the 10 most common NA messages.

```{r}
colorado %>% 
  count(not_available) %>%
  group_by(not_available) %>%
  arrange(-n) %>%
  ungroup %>%
  slice(1:10)
```

Out of 2280 rows, there are `r 2280 - 1792` (2280-1792) rows with some sort of `not_available` message. 

This means that `r 448 / 2280 * 100`% of the data has some sort of `not_available` message to it, which is relatively low.

Most of the data exists, and almost all jails at least provide some sort of ethnicity data. Most of the data that is missing is that for specific measures as aforementioned above, a jail's JMS (Jail Management System) might not break down types of sentences by gender, race, or ethnicity. When conducting data analysis on race and gender for some particular measures, it will be a good idea to remove these rows, or at least account for them. 

# 2. Check Class

Check class of data:

```{r glimpse at the data}
glimpse(colorado)
```
Judging from the datatypes above, we will change Qtr to become a factor variable, because Qtr represents periodic stages of data collection, not a continuous value.

```{r}
colorado <- colorado %>%
  mutate(qtr = as.factor(qtr))
```

Overall, the other variables seem to have the correct data type.

# 3. Investigate Missingness

```{r}
colorado <- colorado %>% 
  mutate(isNA = !is.na(not_available)) 

colorado %>%
  count(qtr, isNA) %>%
  filter(isNA == TRUE)

colorado
```

Missingness was generally reduced throughout the three quarters of jail data collection in 2020, possibly suggesting improvements in jail collection throughout this time period.

# 4. EDA

```{r}
demo %>%
  ggplot(aes(ruralurban_cc)) + 
  geom_bar() + 
  labs(title = "There's a normal distribution of counties per class")
```


```{r}
colorado %>%
  count(jms) %>%
  arrange(-n) %>%
  mutate(n = n / 5) %>%
  slice(1:8) %>%
  ggplot(aes(x = reorder(jms, -n), y = n)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Jail Management System",
       y = "Count",
       title = "Top 8 utilized jail management systems in Colorado")
```


```{r}
colorado_population <- colorado %>%
  arrange(qtr) %>%
  group_by(qtr) %>%
  summarise(total = sum(total)) 

colorado_population
```

As you can see, Colorado jail population has significantly reduced during COVID-19. 

```{r Jail Population Distribution}
# facet version

colorado_dist <- colorado %>%
  group_by(county, qtr) %>%
  summarise(total = sum(total))

colorado_dist %>%
  ggplot(aes(x = total)) + 
  geom_histogram(bins = 50) + 
  facet_grid(qtr ~ .) + 
  labs(title = "Distribution of Colorado Jail Size (by Population) in different
Quarters",
       y = "Count",
       x = "Jail Population")
```

The population per jail throughout Colorado is Unimodal, right-skewed distribution with significant outliers on the right of the
graph.

```{r List the Jails with the Largest Populations}
colorado_dist %>%
  filter(total > 20000) %>%
  distinct(county)
```

The largest jails in Colorado are the Adams, Arapahoe, and El Paso County Jails.

```{r Distribution Statistics}
total_summary <- colorado_dist %>%
  summarise(mean_total = mean(total),
    sd_total = sd(total),
    median_total = median(total),
    IQR_total = IQR(total))

total_summary
```

# 5. Linear Regression

```{r}
colorado_num_inmates <- colorado %>%
  filter(measure == "Number of inmates")

colorado_num_inmates %>%
  filter(county == "Pitkin" | county == "Pueblo")
```

Since only two jails (Pitkin and Pueblo) don't offer information on ethnicity or race for the measure of number of total inmates in only one of their quarters, this data, focused on ethnicity counts per county, is mostly complete for modelling. We'll remove these numbers

Let's focus on only quarter 1 and 3:

```{r}
colorado_num_inmates <- colorado %>%
  filter(measure == "Number of inmates") %>%
  filter(qtr == 1 | qtr == 3)

colorado_num_inmates
```

```{r}
colorado_num_inmates %>%
  count(county) %>%
  filter(n == 1)
```
Remove 4 jails that don't have both first and third quarter: Grand, Huerfano, Las Animas, Saguache

```{r}
colorado_num_inmates <- colorado_num_inmates %>%
  filter(county != "Grand" & 
           county != "Huerfano" &
           county != "Las Animas" &
           county != "Saguache") %>%
  select(-c(not_available, isNA, jms, qtr_year, measure, deaths, 
            other_gender, 
            bookings, releases)) %>%
  mutate(other_race = unknown_race + other_race) %>%
  select(-c(unknown_race)) %>%
  arrange(county)

colorado_num_inmates
```


Pivot_wider;

```{r pivot wider}
colorado_num_inmates <- colorado_num_inmates %>% 
  pivot_wider(names_from = qtr, values_from = capacity:unknown_ethnicity) 
```

```{r join the datasets}
colorado_num_inmates <- colorado_num_inmates %>%
  inner_join(pop, by = "county") %>%
  inner_join(demo, by = "county")
```

```{r}
colorado_num_inmates <- colorado_num_inmates %>%
  mutate(difference = (total_3 - total_1)/total_1)

colorado_num_inmates
```

```{r standardizing pcts}
colorado_num_percent <- colorado_num_inmates %>%
  mutate(jail_male_pct = male_1 / total_1) %>%
  mutate(jail_black_pct = black_1 / total_1) %>%
  mutate(jail_hispanic_pct = hispanic_1 / total_1) %>%
  select(county, difference, lesscollege_whites_pct, jail_male_pct, jail_black_pct, jail_hispanic_pct,
         population, rural_pct, black_pct, liberal, ruralurban_cc)

colorado_num_percent
```

# new model 

```{r model fit}
full <- lm(difference ~ liberal + 
     lesscollege_whites_pct + 
     population +
     jail_male_pct + 
     jail_black_pct + 
     jail_hispanic_pct+ 
     ruralurban_cc + 
     black_pct,
   data = colorado_num_percent)

full %>%tidy
```

```{r backwards selection}
int_only_model <- lm(difference ~ 1, data = colorado_num_percent)

covid_model <- step(full, scope = formula(int_only_model), direction = "backward")
```

```{r}
covid_model %>%
  tidy(conf.int = TRUE)
```

Based on backwards AIC selection, the two significant predictors for are the percent of males in a jail and whether a jail is in a rural, urban, or metropolitan area.

A jail in an urban or metropolitan area is expected to decrease 1 percent more than a jail in a rural area, on average.

For every one percent increase in male inmate in a jail, there is expected to be a 5 percent additional decrease in jail population, on average.



## Interaction Term

```{r nested F-test}
reduced_model <- final_model
full_model <- lm(unemployment_change ~ retail_avg + grocery_avg + workplaces_avg + retail_avg*grocery_avg, data = sample_data)

anova(reduced_model, full_model) %>%
  tidy() %>%
  kable(digits = 3)
```

## Model Conditions

```{r conditions}
model_aug <- augment(full_model) %>%
   mutate(obs_num = row_number()) #add row number to help with graphing

resid_fitted <- ggplot(data = model_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted values",
    y = "Residual",
    title = "Residuals vs. Predicted") +
  theme_ipsum_rc()

resid_hist <- ggplot(data = model_aug, aes(x = .resid)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Residuals", title = "Dist. of Residuals") +
  theme_ipsum_rc()

resid_qq <- ggplot(data = model_aug, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal QQ-plot of residuals") +
  theme_ipsum_rc()

conditions_plot <- resid_fitted / (resid_hist + resid_qq)
```

```{r save file}
ggsave(here("output", "test-plot.png"), conditions_plot, units = "in", width = 850/100, height = 600/100)
```

```{r}
#test_plot <- ggplot(mtcars, aes(mpg, wt)) +
  #geom_point() +
  #labs(x="Fuel efficiency (mpg)", y="Weight (tons)",
       #title="Seminal ggplot2 scatterplot example",
       #subtitle="A plot that is only useful for demonstration purposes",
       #caption="Brought to you by the letter 'g'") +
  #theme_ipsum_rc()
```

```{r save file}
#ggsave("../output/test-plot.png", test_plot, units = "in", width = 850/100, height = 600/100 )
```



###


## Model Diagnostics

**Leverage**

```{r leverage threshold}
#calculate threshold
leverage_threshold <- 2*(3+1)/500
leverage_threshold
```

```{r leverage v obs num}
ggplot(data = model_aug, aes(x = obs_num, y = .hat)) + 
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept = leverage_threshold, color = "red") +
  labs(x = "Observation Number", y = "Leverage") +
  geom_text(aes(label=ifelse(.hat > leverage_threshold,
                             as.character(obs_num), "")), nudge_x = 4)
```

```{r obs w high leverage}
model_aug %>% 
  filter(.hat > leverage_threshold)
```

**Standardized residuals**

```{r std resid}
#scatterplot of std resid vs predicted
ggplot(data = model_aug, aes(x = obs_num, y = .std.resid)) + 
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept = 0, color = "red") +
  geom_hline(linetype = "dotted", yintercept = c(-2,2)) +
  geom_hline(linetype = "dotted", yintercept = c(-3,3), color = "red") +
  labs(x = "Observation Number", y = "Standardized Residuals") +
  geom_text(aes(label=ifelse(.std.resid > 3,
                             as.character(obs_num), "")), nudge_x = 4)
```

```{r potential outliers}
model_aug %>% 
  filter(abs(.std.resid) > 3)
```

**Cook's distance**

```{r cooks d}
#scatterplot of cook's d vs obs num
ggplot(data = model_aug, aes(x = obs_num, y = .cooksd)) + 
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept = 1, color = "red") +
  geom_hline(linetype = "dotted", yintercept = 0.5, color = "red") +
  labs(x = "Observation Number", y = "Cook's D") +
  geom_text(aes(label=ifelse(.cooksd > 1,
                             as.character(obs_num), "")), nudge_x = 4)
```

**Multicollinearity**

```{r check multicollinearity}
vif(full_model) %>%
  tidy() %>%
  kable(digits = 3)
```

All of the predictor variables have a VIF less than 10. Thus, we can say that none of the predictor variables in our model are correlated.

## Full model with interaction term

$$unemployment\_change = 1.266 - 0.73\times{retail\_avg} - 0.015\times{grocery\_avg} -0.035\times{workplaces\_avg} + 0.002\times{retail\_avg:grocery\_avg}$$

```{r}
tidy(full_model, conf.int = TRUE) %>% 
  kable(digits = 3)
```

## Model comparison

```{r}
tidy(final_model) %>% 
  kable(digits = 3)
```

```{r}
glance(final_model) %>% 
  select(AIC,BIC,r.squared,adj.r.squared)
```

```{r}
tidy(full_model) %>% 
  kable(digits = 3)
``` 

```{r}
glance(full_model) %>%
  select(AIC,BIC,r.squared,adj.r.squared)
```

```{r}
model2 <- lm(unemployment_change ~ retail_avg*grocery_avg, data = sample_data)
tidy(model2, conf.int = TRUE) %>% 
  kable(digits = 3)
```

```{r}
glance(model2) %>% 
  select(AIC,BIC,r.squared,adj.r.squared)
```

## Interpretation (Edit)

```{r}
tidy(full_model) %>% 
  kable(digits = 3)
``` 

The intercept is 28.454. This means that the expected median property value of owner-occupied homes for towns located in the suburbs of Boston with `r round(mean(real_estate$CRIM))` crime rate per capita, no residential land zoned for lots over 25,000 sq.ft, `r round(mean(real_estate$INDUS))`, does not bound river, with `r round(mean(real_estate$RM))` number of rooms per dwelling on average, no owner-occupied units built prior to 1940, `r round(mean(real_estate$DIS))` as weighted distance to five Boston employment centers, and pupil-teacher ratio of `r round(mean(real_estate$PTRATIO))` is around 28454 dollars.

The estimated coefficient for `indus_cent:CHAS1` is 0.936. This means that for every one percentage point increase in the proportion of non-retail business acres per town, the median property value of owner-occupied homes for each town is expected to increase by 936 dollars, on average, for towns the bound the Charles River than towns that do not bound the Charles River, holding all other predictor variables constant.

Based on our current output, we can see that both geographical and societal factors influence the property values in Boston suburbs. In particular, factors that indicate better environment and number of rooms in a house have a positive linear relationship with the property values in Boston suburbs. On the other hand, proportion of non-retail business, proportion of old buildings, distances to employment centers, and pupil-teacher ratio have negative linear relationships with the property values in Boston suburbs. This is roughly in line with what we expected since characteristics that indicate limited access to resources (e.g., high student-to-teacher ratio andd low average number of rooms per home) tend to have have lower median property values of owner-occupied homes.


###

Comments: 

- didn't check model conditions yet

- LOOK AT DISCHARGE

- This doesn't look at the "after" numbers for any of the predictor variables (race, gender, ethnicity). Another thing we can do in the future is analyze the difference between races over time.

- are the race variables independent enough? 

- try looking at which race percentage decreases the most

- look at bed increases or decreases

- look at capacity changes

- make helper functions to streamline the creation of other regression models

Ideas:

RQ: Is there a difference between the proportion of people released from COVID-19 between races? Gender?

- ANova means to asess whether there's been a significant change due to COVID-19

- Look at Legalization, COVID-19, COVID-19 on Race

- How do I research whether having a larger percentage of Black people made jails more/less likely to release people?

- Add judiciary political, national political, county political (assembly member elected)

- ggmap