---
title: "HB 19-1297: COVID-19 Intake vs. Outtake"
output: pdf_document
---

```{r message=FALSE}
library(tidyverse)
library(patchwork)
library(janitor)
library(broom)
library(knitr)
library(here)
```

```{r global options, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r read data}
colorado <- read_csv(here("data", "HB19-1297Data.csv")) %>%
  clean_names()

pop <- read_csv(here("data", "colorado-population.csv")) %>%
  clean_names()

pop <- pop %>%
  filter(county != "Total") %>%
  mutate(county = str_sub(county, 2, -18))
```

```{r}
colorado %>%
  filter(measure == "Number of inmates") %>%
  filter(qtr == 1 | qtr == 3)
```


# Pivot

Pivot the table to create `colorado_change`, a horizontal table that where a county observation has these counts for each period: # total, deaths, bookings, releases, male, female, other gender, black, native american, other race, white, unknown race, hispanic/not/unknown ethnicity.

```{r change betwen quarter 1 and 3}
colorado_change <- colorado %>%
  filter(measure == "Number of inmates") %>%
  pivot_wider(id_cols = c(-not_available), names_from = qtr, values_from = c(total, deaths:releases, total:unknown_ethnicity)) 

colorado_change
```

We're going to focus on the net population change and the total bookings and releases during this 2 quarter period (Janurary to March vs. March to June) of bookings and releases over time. 

```{r}
colorado_change <- colorado_change %>%
  mutate(population_change = total_1 - total_2) %>%
  mutate(bookings_change = bookings_1 - bookings_2) %>%
  mutate(releases_change = releases_1 - releases_2) %>%
  select(county, population_change, bookings_change, releases_change)
```

Add total population counts and Rural, UA, and UC

ADD WHAT THESE MEAN

```{r add total population counts}
colorado_change <- inner_join(colorado_change, pop, by = c("county" = "county")) %>%
  mutate(type = if_else(population < 2500, "R", 
                        if_else(population < 50000, "UC", "UA")))

colorado_change 
```

```{r}
colorado_change %>% 
  summarise(non_na_count = sum(!is.na(population_change)))
```

Only 33 of the counties have their intake and outtake counts.

INVESTIGATE WHICH COUNTIES THESE ARE

```{r}
colorado_change <- colorado_change %>%
  drop_na()

colorado_change
```


# Model

```{r}
intake_model <- lm(population_change ~ bookings_change, data = colorado_change)

intake_model %>%
  tidy()
```

```{r}
outtake_model <- lm(population_change ~ releases_change, data = colorado_change)

outtake_model %>%
  tidy()
```

It's clear that reduced bookings has a much more significant effect on population change than increased releases do. 

Although the media in Colorado tend to highlight on increased releases, they still factor into population change less than releases do. 

DO NEXT: 

- check region variables

- change population type variables