---
title: "second-sem-EDA"
author: "Albert Sun"
date: "2/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(patchwork)
library(janitor)
library(broom)
library(knitr)
library(here)
library(rms)
library(skimr)
```

## Plot

```{r}
colorado <- read_csv(here("data", "HB19-1297Data.csv")) %>%
  clean_names()

demo <- read_csv(here("data", "mit-demographics.csv")) %>%
  filter(state == "Colorado") 
```

```{r}
demo <- demo %>%
  mutate(liberal = factor(if_else(clinton16 - trump16 > 0, 1, 0))) %>%
  mutate(urbanicity = factor(ruralurban_cc)) %>%
  mutate(urbanicity = fct_collapse(urbanicity, 
                      metro = c("1", "2", "3"), 
                      urban = c("4", "5", "6", "7"), 
                      rural = c("8", "9"))) %>%
  mutate(urbanicity = fct_relevel(urbanicity,
                                     'rural',
                                     'urban'))

demo
```

```{r}
colorado_num_inmates <- colorado %>% 
  filter(measure=='Number of inmates') %>%
  group_by(county, qtr) %>%
  arrange(county, qtr)
```

```{r}
how_urban <- demo %>% select(county, urbanicity)
colorado_num_inmates <- merge(colorado_num_inmates, how_urban, by='county')
colorado_num_inmates
```
```{r}
urbanicity_df <- colorado_num_inmates %>%
  group_by(urbanicity, qtr) %>%
  summarise(total_population = sum(total))

urbanicity_df
```



```{r}
quarter_labs <- c("Q1 (Jan 1, 2020)", "Q2 (Apr 1, 2020)", "Q3 (Jul 2020)", "Q4 (Oct 2020)")
date_labs <- c("1/1/2020", "4/1/2020", "7/1/2020", "10/1/2020")

#urbanicity_df$qtr <- as.factor(urbanicity_df$qtr)
urbanicity_df$urbanicity <- factor(urbanicity_df$urbanicity, levels = c("metro", "urban", "rural"))

ggplot(data = urbanicity_df, aes(x = factor(qtr), y = total_population, color = urbanicity, group = urbanicity)) +
  geom_line(size=2) + 
  ggtitle("Jail Population by Urbanicity in Colorado") + 
  scale_x_discrete(labels=date_labs) +
  theme_gray() + 
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank()) 
  
```

Use this urbanicity
https://www.colorado.gov/pacific/sites/default/files/DC_STI_HIVPrev_Colorado-County-Designations.pdf:

```{r}
designations <- read_csv(here("data", "colorado_designations.csv")) 

designations
```

```{r}
colorado %>% 
  filter(measure == 'Number of inmates') %>%
  select(county, qtr, total) %>%
  left_join(designations, by = c("county" = "County")) %>%
  group_by(Urbanicity, qtr) %>%
  summarise(total_population = sum(total)) %>%
  ggplot(aes(x = qtr, y = total_population, color = Urbanicity)) +
  geom_line()
```

Rural counties and frontier counties faced a smaller percentage decrease at the beginning of COVID-19, suggesting that urban jails made a bigger, deliberate effort to de-densify jails than suburban jails. 

General Summary Statistics:

```{r}
colorado_num_percent %>%
  ggplot(aes(x = difference)) + 
  geom_histogram(bins = 30) + 
  labs(title = "Distribution of population percent changes during COVID-19",
       x = "% Difference",
       y = "Count"
         )
```

The percent change in population numbers has a unimodal normal distribution. It has a slight right-skew with one significant outlier at +7.5% difference.

```{r}
demo %>%
  mutate(urbanicity = fct_relevel(urbanicity, "urban")) %>%
  ggplot(aes(urbanicity)) + 
  geom_bar() + 
  labs(title = "Distribution of counties per urbanicity",
       y = "Count", 
       x = "Urbanicity")
```

```{r}
colorado %>%
  mutate(jms = str_replace(jms, "Eforce", "E-Force")) %>%
  mutate(jms = str_replace(jms, "eForce", "E-Force")) %>%
  count(jms) %>%
  arrange(-n) %>%
  mutate(n = n / 5) %>%
  slice(1:5) %>%
  ggplot(aes(x = reorder(jms, -n), y = n)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Jail Management System",
       y = "Count",
       title = "Top 5 utilized jail management systems in Colorado")
```

The top utilized jail management system in Colorado are E-Force, Spillman, New World, Crimestar, and Tiburon.

```{r}
colorado %>%
  arrange(qtr) %>%
  group_by(qtr) %>%
  summarise(total = sum(total)) %>%
  ggplot(aes(x = qtr, y = total)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Total Population in Colorado Jails at first three Quarters of 2020",
       x = "Quarter",
       y = "Population #")
```

As you can see, Colorado jail population has significantly reduced during COVID-19. 

```{r Jail Population Distribution}
# facet version
colorado_dist <- colorado %>%
  group_by(county, qtr) %>%
  summarise(total = sum(total))

colorado_dist %>%
  ggplot(aes(x = total)) + 
  geom_histogram(bins = 50) + 
  facet_grid(qtr ~ .) + 
  labs(title = "Distribution of Colorado Jail Size (by Population) in different
Quarters",
       y = "Count",
       x = "Jail Population")
```

The population per jail throughout Colorado is Unimodal, right-skewed distribution with significant outliers on the right of the
graph.